Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-29 paper2.qmd
section:6; process:write

[^1]: These topics are beyond the scope of this study, thus, readers seeking a more profound understanding can refer to introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020, and introductory books like @Pearl_et_al_2018, @Neal_2020, and @McElreath_2020 are useful. For more advanced study, seminal papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, along with books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2020, are recommended.

@@

<!-- 2. what is the data structure? -->
Here, each element of $I$ represents a unique individual $i$ or $h$, where $n_{I}$ denotes the total number of individuals. Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges. Moreover, each row of $IA$ represents a unique pairing of individuals $i, h$ with stimuli $a, b$. As a result, the matrix $IA$ contains $n_{I} \cdot n_{A}$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual. Likewise, each row of $JK$ associates a judge $j$ with a (repeated) judgment index $k$. Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of repeated judgments each judge makes.

Additionally, we construct the matrix $R$ to map each row of the $IA$ matrix with a corresponding row from the $JK$ matrix. This matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$. Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population. Thus, we define the matrix as follows:

$$
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
i & a & h & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}
\end{bmatrix}
$$ {#eq-mat12}

It is easier to visualize the structure of these vectors and matrices by considering an example. Assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the vectors and matrices described in Equations ([-@eq-mat11]) and ([-@eq-mat12]) take the following form:
$$
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 2 & 1 & 1 \\
1 & 1 & 5 & 2 & 1 & 2 \\
1 & 1 & 5 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
4 & 2 & 5 & 2 & 3 & 1 \\
4 & 2 & 5 & 2 & 3 & 2 \\
4 & 2 & 5 & 2 & 3 & 3 \\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 
\end{bmatrix}
$$ {#eq-mat13}

@@

Finally, we define the *sample mechanism* $S$ in Equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$. Each element $s_{(i,a,h,b,j,k)}$ is a binary value indicating the presence or absence of data rows in the matrix $R$ resulting from the sample mechanism, as in Equation ([-@eq-mat26]). Thus, the vector contains $n$ elements, matching the number of rows in $R$, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$. Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.

$$
S = \begin{bmatrix}
s_{(1,1,1,2,1,1)} \\
\vdots \\
s_{(1,1,1,2,1,n_{K})} \\
\vdots \\
s_{(i,a,h,b,j,k)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
$$ {#eq-mat25}

@@

[^8]: Randomization ensures that data--and, by extension, an estimator--satisfies several key identification properties, such as common support, no interference, and consistency. The most critical property, however, is the elimination of confounding. *Confounding* occurs when an external variable, such as $X_{I}$, simultaneously influences both the outcome (e.g., $O_{R}$) and a variable of interest (e.g., $S$), resulting in spurious associations between the latter two [@Everitt_et_al_2010]. Randomization ensure the absence of confounding by effectively decoupling the association between the variable of interest and any other variable, except for the outcome itself. For a more detailed discussion on the benefits of randomization, see @Pearl_2009, @Morgan_et_al_2014, @Neal_2020, and @Hernan_et_al_2020.

@@

<!-- 1. Summary of Thurstone's general theory -->
Thurstone introduced the Law of Comparative Judgment to measure psychological traits of stimuli through pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. In its general form, the theory models single-judge comparisons across multiple, potentially correlated stimuli. Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level. However, Thurstone identified one key challenge with its measurement model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. To address this issue and facilitate the theory's practical applicability, he formulated five cases derived from this general form, each progressively incorporating several simplifying assumptions.

<!-- 2. Summary of the why the literature uses Case V, and its problems -->
Of these five cases, the CJ literature has predominantly relied on Case V to measure various psychological traits, as shown in studies like @Kimbell_2012, @Jones_et_al_2015, and @Boonen_et_al_2020. This preference largely stems from the widespread adoption of the BTL model, which offers a simplified statistical formulation of the case. The BTL model mirrors the assumptions of Case V--namely, equal discriminal dispersions and zero correlation for the stimuli' discriminal processes--with one notable distinction. While Case V assumes normally distributed discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008]. Although this substitution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], Thurstone acknowledged that assuming equal discriminal dispersions and zero correlation among stimuli' discriminal processes could introduce significant problems. In particular, he recognized that while these assumptions simplify the trait measurement model, they may not capture the complexity of some traits or account for heterogeneous stimuli, such as handwritten texts or English compositions [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022]. As a result, they can lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].

Furthermore, because Thurstone aimed primarily to produce a trait's "coarse scaling" and allocate the compared stimuli along this continuum [@Thurstone_1927a, pp. 269], his theory did not specify how to use the resulting trait estimates for statistical inference. The CJ tradition has attempted to address this gap by separating trait estimation from hypothesis testing. Specifically, CJ studies often rely on point estimates of the traits--typically the BTL scores or its transformations--to conduct statistical inference. Although this separation simplifies CJ data analysis, the statistical literature warns that this practice can introduce bias into the analysis and compromise the reliability of statistical inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

<!-- 4. General summary of what we have done. -->
Thus, to address the limitations of Thurstone's Case V and the BTL model--particularly their strong assumptions and the disconnect between trait measurement and hypothesis testing--this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods. The approach begins by formulating a conceptual model represented by a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG). This model integrates Thurstone's core theoretical principles--such as the discriminal processes of stimuli--alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes in the CJ system.

@@

Sampling and comparison mechanisms play a central role in modern CJ experimental design. However, most of the CJ literature has examined these mechanisms within a limited scope. Researchers have primarily investigated the effects of adaptive comparative judgment (ACJ) designs on trait reliability [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021; @Gray_et_al_2024] or proposed practical guidelines for the number of comparisons judges should make [@Verhavert_et_al_2019; @Crompvoets_et_al_2022]. Although these studies offer valuable insights, we argue that they may have overlooked the broader role these mechanisms play within the CJ system. Moreover, we suggest that this oversight largely stems from a lack of conceptual clarity about how the mechanisms function within the system. To address this gap, this study decided to integrate these mechanisms into CJ's conceptual model.

The explicit integration of the sampling and comparison mechanisms offers a new perspective on how these mechanisms shape the CJ process. In particular, it clarifies their role in the data-generating process by framing them as sources of missing data--that is,  as mechanisms that determine which observations are missing from the final data sample. This new perspective encourages the application of Little and Rubin's principled missing data framework [-@Little_et_al_2020], enabling a more rigorous evaluation of existing claims about these mechanisms, their influence on CJ outcomes, and their implications for designing and assessing more complex experimental setups.

Notably, this study circumvents the need to apply this missing data framework by deliberately structuring the sampling and comparison mechanisms to be independent of any observed or unobserved variables, including the outcome. In other words, we designed these mechanisms so they produce data that are *missing completely at random* (MCAR) [@Little_et_al_2020]. This design offers one key advantage: it generates simple random samples that satisfy the condition of *ignorability*, allowing researchers to legitimately *ignore* missing data during analysis without introducing bias [@Everitt_et_al_2010; @Kohler_et_al_2019; @Neal_2020].

@@

However, many modern CJ applications rely on more complex experimental designs in which the sampling and comparison mechanisms introduce more intricate forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020]. A prominent example is the previously discussed ACJ designs, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015]. This outcome-informed selection suggests that the comparison mechanism may be outcome-dependent, potentially classifying ACJ as a generator of MNAR data. This classification may, in turn, facilitate a more nuanced understanding of the mixed evidence surrounding ACJ's capabilities. While some studies report improvements in trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b],  others argue that these gains are artificially inflated by the method [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].

@@

Nevertheless, regardless of the underlying missingness mechanisms, any CJ experimental design would benefit from explicitly defining its assumptions--an approach supported by our framework. This clarity enables researchers to evaluate, for instance, how sampling and comparison mechanisms affect trait estimation and statistical inference within each design. Such assessments are particularly relevant given the common misconception in the CJ literature that Thurstone's model can naturally handle even non-random missing data without compromising the reliability or validity of trait estimates [@Bramley_2008].

@@

Despite the growing notion that various stimulus-related factors can influence judges' perceptions during pairwise comparisons--and that these influences may not always cancel each other out--few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a]. We argue that this gap persists not due to a lack of interest or research but because researchers continue to rely on ad-hoc detection methods, such as 'misfit' statistics, that are poorly suited for the task [@Kelly_et_al_2022]. To address this limitation, this study decided to treat judges' biases as an integral component of the CJ system from the outset. This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons--one that acknowledges that the discriminal processes of stimuli become an observable outcome only through judges' perceptions, which may exhibit bias.

The explicit integration of judges' bias into CJ's conceptual model then paves the way for investigating several relevant research questions. For instance, can researchers validly analyze CJ data under the assumption of "sample-free" trait calibration--specifically, under the assumption that judges exhibit no systematic bias? This question is especially relevant because researchers often treat "sample-freeness" as an inherent property of the BTL model [@Bramley_2008; @Andrich_1978] despite growing evidence of persistent biases. Furthermore, if judges' biases reflect additional stimulus characteristics that should not influence assessment, to what extent can training or expertise mitigate these biases? [@Kelly_et_al_2022] More specifically, as discussed in @sec-theory-issue1b, can training or expertise help prevent judges from prioritizing irrelevant features--such as handwriting--over more central criteria like argumentative quality in writing assessments? Addressing these questions may also shed light on what it truly means to be an "expert" in the CJ context [@Kelly_et_al_2022].

@@

Moreover, if we accept that judges may value different aspects of a stimulus, it follows that their judgments might also vary with individual characteristics such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022; @Bartholomew_et_al_2020a], and even from one individual to another [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020]. Building on this and the discussion in @sec-discussion_RA1, researchers could investigate how the selection of judges influences the formation of a "shared consensus" and whether these individual characteristics introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022]. Furthermore, if such characteristics indeed undermine the "sample-freeness" assumption, then it becomes essential to explore strategies for controlling their influence. This includes investigating how many judges--and how many judgments per judge--are required to produce reliable trait estimates, as well as whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.

@@

<!-- 7.1 PPL and Bayesian Inference -->
<!-- Despite its potential, this study presents several challenges that researchers must navigate. First, researchers need to be proficient in Probabilistic Programming Languages (PPLs) and Bayesian inference methods. This expertise is essential for translating context-specific conceptual models into bespoke statistical models that meaningfully address the research questions at hand. While such translations are not exclusive to Bayesian methods, we argue that Bayesian approaches offer a more accessible path. They reduce, for instance, the need for additional expertise in optimization techniques and theory often required by frequentist methods, thereby lowering the technical barrier for many applied researchers. -->

<!-- make these four into one general challenge, "something you can always say about of almost every study". Now the challenge need a thorough knowledge about the CJ system. No easy route now, they have to think about their research context, assumptions in the experiment, model. -->

<!-- our framework is not designed to work for everything, is just a starting point to think about which model applies to your research context  -->

<!-- "although in this paper we have outlined the causal process behind the CJ system, this does not refrain other research to thoroughly think their causal structure in their studies" -->

<!-- 7.2 Research context is important  -->
<!-- We have argued that the model's ability to accurately estimate traits and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions. That said, our approach provides tools for researchers to test and evaluate these context-dependent CJ models directly. -->

<!-- 7.4 Causal inference knowledge -->
<!-- Fourth, applying this approach requires a solid understanding of Causal inference. For instance, throughout the study, we assume that predictor variables must be "relevant" to the research context, interpreting relevance as the inclusion of the predictor variables in a "sufficient adjustment set." However, we do not fully explore what this entails or its implications for model specification and validity. Nonetheless, we provide key references throughout the study to help motivated researchers explore these foundational concepts in greater depth. -->

<!-- 7.5 Epistemiology  -->
<!-- Fifth and final, our approach does not establish a clear epistemological connection between the theoretical latent trait and its quantitative estimate. In other words, the approach does not provide a definitive answer to the fundamental question: How do we know that the estimated trait reflects the theoretical trait of interest? However, this limitation is not unique to our approach; it reflects a fundamental challenge all research involving latent constructs faces. Nevertheless, we argue that our approach offers methodological tools that enable researchers to design CJ experiments under conditions that support reliable and valid trait measurement. -->

<!-- 7.3 Faithfulness  -->

<!-- remember to mention fidelity of the theory in the model -->

>>>>

[^1]: These topics are beyond the scope of this study, thus, readers seeking a more profound understanding can refer to introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020, and introductory books like @Pearl_et_al_2018, @Neal_2020, and @McElreath_2020 are useful. For more advanced study, seminal papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, along with books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2025, are recommended.

@@

<!-- 2. what is the data structure? -->
Here, each element of $I$ represents a unique individual $i$ or $h$, where $n_{I}$ denotes the total number of individuals. Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges. Moreover, each row of $IA$ represents a unique pairing of individuals $i, h$ with stimuli $a, b$. As a result, the matrix $IA$ contains $n_{I} \cdot n_{A}$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual. Likewise, each row of $JK$ associates a judge $j$ with a (repeated) judgment index $k$. Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of repeated judgments each judge makes.
$$
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
i & a & h & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}
\end{bmatrix}
$$ {#eq-mat12}

Additionally, we construct the matrix $R$ to map each row of the $IA$ matrix with a corresponding row from the $JK$ matrix. This matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$. Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population. Thus, we define the matrix as in Equation ([-@eq-mat12]).

It is easier to visualize the structure of the previously defined vectors and matrices by considering an example. Assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the vectors and matrices described in Equations ([-@eq-mat11]) and ([-@eq-mat12]) take the form as in Equation ([-@eq-mat13]).
$$
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 2 & 1 & 1 \\
1 & 1 & 5 & 2 & 1 & 2 \\
1 & 1 & 5 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
4 & 2 & 5 & 2 & 3 & 1 \\
4 & 2 & 5 & 2 & 3 & 2 \\
4 & 2 & 5 & 2 & 3 & 3 \\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 
\end{bmatrix}
$$ {#eq-mat13}

@@

Finally, we define the *sample mechanism* $S$ in Equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$. Each element $s_{(i,a,h,b,j,k)}$ is a binary value indicating the presence or absence of data rows in the matrix $R$ resulting from the sample mechanism, as in Equation ([-@eq-mat26]). Thus, the vector contains $n$ elements, matching the number of rows in $R$, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$. Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.
$$
S = \begin{bmatrix}
s_{(1,1,1,2,1,1)} \\
\vdots \\
s_{(1,1,1,2,1,n_{K})} \\
\vdots \\
s_{(i,a,h,b,j,k)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
$$ {#eq-mat25}

@@

[^8]: Randomization ensures that data--and, by extension, an estimator--satisfies several key identification properties, such as common support, no interference, and consistency. The most critical property, however, is the elimination of confounding. *Confounding* occurs when an external variable, such as $X_{I}$, simultaneously influences both the outcome (e.g., $O_{R}$) and a variable of interest (e.g., $S$), resulting in spurious associations between the latter two [@Everitt_et_al_2010]. Randomization ensure the absence of confounding by effectively decoupling the association between the variable of interest and any other variable, except for the outcome itself. For a more detailed discussion on the benefits of randomization, see @Pearl_2009, @Morgan_et_al_2014, @Neal_2020, and @Hernan_et_al_2025.

@@

<!-- 1. Summary of Thurstone's general theory -->
Thurstone introduced the Law of Comparative Judgment to measure psychological traits of stimuli through pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. In its general form, the theory models single-judge comparisons across multiple, potentially correlated stimuli. Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level. However, Thurstone identified one key challenge with its measurement model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. To address this issue and facilitate the theory's practical applicability, he formulated five cases, each progressively incorporating several simplifying assumptions.

<!-- 2. Summary of the why the literature uses Case V, and its problems -->
Among these, Case V has become the most widely used model in empirical CJ research, primarily due to the widespread adoption of the BTL model. The BTL model mirrors the core assumptions of Case V--namely, equal discriminal dispersions and zero correlation among stimuli' discriminal processes--but replaces the processes' normal distribution with the more mathematically tractable logistic distribution [@Andrich_1978; @Bramley_2008]. Although the substitution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021] and simplifies the measurement model, the other two core assumptions of the model can still lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023], especially when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a; @Andrich_1978; @Bramley_2008; @Kelly_et_al_2022].

Furthermore, because Thurstone primarily aimed to produce a trait's "coarse scaling" and allocate the compared stimuli along this continuum [@Thurstone_1927a, pp. 269], his theory did not specify how to use the trait estimates for statistical inference. The CJ tradition has attempted to fill this gap by separating trait estimation from hypothesis testing. Specifically, researchers often rely on point estimates of the traits, such as the BTL scores or their transformations, for inference. Although this approach simplifies analysis, the statistical literature cautions that it can introduce bias and compromise the reliability of the resulting inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

<!-- 4. General summary of what we have done. -->
Thus, to address these limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods. The approach begins by formulating a conceptual model represented by a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG). This model integrates Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes in the CJ system.

@@

<!-- 1. The overlooked role of the sampling and comparison mechanisms in CJ -->
Sampling and comparison mechanisms play a central role in modern CJ experimental design. However, most of the CJ literature has examined these mechanisms within a limited scope. Researchers have primarily investigated the effects of adaptive comparative judgment (ACJ) designs on trait reliability [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021; @Gray_et_al_2024] or proposed practical guidelines for the number of comparisons judges should make [@Verhavert_et_al_2019; @Crompvoets_et_al_2022]. Although these studies offer valuable insights, we argue that they may have overlooked the broader role these mechanisms play within the CJ system. Moreover, we suggest that this oversight largely stems from a lack of conceptual clarity about how the mechanisms function within the system. To address this gap, this study decided to integrate these mechanisms into CJ's conceptual model.

<!-- 2. The missing data perspective on sampling and comparison mechanisms -->
The explicit integration of the sampling and comparison mechanisms offers a new perspective on how these mechanisms shape the CJ process. In particular, it clarifies their role in the data-generating process by framing them as sources of missing data--that is, as mechanisms that determine which observations are missing from the final data sample. This new perspective encourages the application of Little and Rubin's principled missing data framework [-@Little_et_al_2020], enabling a more rigorous evaluation of existing claims about these mechanisms, their influence on CJ outcomes, and their implications for designing and assessing more complex experimental setups.

<!-- 3. Leveraging MCAR assumptions to simplify CJ analysis in this study -->
Notably, this study circumvents the need to apply this missing data framework by deliberately structuring the sampling and comparison mechanisms to be independent of any observed or unobserved variables, including the outcome. In other words, we designed these mechanisms so they produce data that are *missing completely at random* (MCAR) [@Little_et_al_2020]. This design offers one key advantage: it generates simple random samples that satisfy the condition of *ignorability*, allowing researchers to legitimately *ignore* missing data during analysis without introducing bias [@Everitt_et_al_2010; @Kohler_et_al_2019; @Neal_2020].

@@

<!-- 4. More complex CJ designs and the implications of MAR or MNAR -->
However, many modern CJ applications rely on more complex experimental designs in which the sampling and comparison mechanisms introduce more intricate forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) [@Little_et_al_2020]. A prominent example is the previously discussed ACJ designs, where prior judgment outcomes inform the selection of stimulus pairs for subsequent comparisons [@Pollitt_2012a; @Pollitt_2012b; @Bramley_2015]. This outcome-informed selection suggests that the comparison mechanism may be outcome-dependent, potentially classifying ACJ as a generator of MNAR data. This classification may, in turn, facilitate a more nuanced understanding of the mixed evidence surrounding ACJ's capabilities. While some studies report improvements in trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b],  others argue that these gains are artificially inflated by the method [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022].

@@

<!-- 5. The importance of clarifying assumptions behind CJ experimental designs -->
Nevertheless, regardless of the underlying missingness mechanisms, any CJ experimental design would benefit from explicitly defining its assumptions--an approach supported by our framework. This clarity enables researchers to evaluate, for instance, how sampling and comparison mechanisms affect trait estimation and statistical inference within each design. Such assessments are particularly relevant given the common misconception in the CJ literature that Thurstone's model can naturally handle even non-random missing data without compromising the reliability or validity of trait estimates [@Bramley_2008].

@@

<!-- 1. Toward a more realistic CJ model: recognizing judges' bias -->
Despite the growing notion that various stimulus-related factors can influence judges' perceptions during pairwise comparisons--and that these influences may not always cancel each other out--few studies in the CJ literature provide empirical evidence for judges' biases [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a]. We argue that this gap persists not due to a lack of interest or research but because researchers continue to rely on ad-hoc detection methods, such as 'misfit' statistics, that are poorly suited for the task [@Kelly_et_al_2022]. To address this limitation, this study decided to treat judges' biases as an integral component of the CJ system from the outset. This approach offers one key advantage: it provides a more accurate representation of the data-generating process behind pairwise comparisons--one that acknowledges that the discriminal processes of stimuli become an observable outcome only through judges' perceptions, which may exhibit bias.

<!-- 2. Relevant research question paved by recognizing judges' bias (part I) -->
The explicit integration of judges' bias into CJ's conceptual model then paves the way for investigating several relevant research questions. For instance, can researchers validly analyze CJ data under the assumption of "sample-free" trait calibration--specifically, under the assumption that judges exhibit no systematic bias? This question is especially relevant because researchers often treat "sample-freeness" as an inherent property of the BTL model [@Bramley_2008; @Andrich_1978] despite growing evidence of persistent biases. Furthermore, if judges' biases reflect additional stimulus characteristics that should not influence assessment, to what extent can training or expertise mitigate these biases? [@Kelly_et_al_2022] More specifically, as discussed in @sec-theory-issue1b, can training or expertise help prevent judges from prioritizing irrelevant features--such as handwriting--over more central criteria like argumentative quality in writing assessments? Addressing these questions may also shed light on what it truly means to be an "expert" in the CJ context [@Kelly_et_al_2022].

@@

<!-- 2. Relevant research question paved by recognizing judges' bias (part II) -->
Moreover, if we accept that judges may value different aspects of a stimulus, it follows that their judgments might also vary with individual characteristics such as gender, age, culture, income, education, training, or expertise [@Kelly_et_al_2022; @Bartholomew_et_al_2020a], and even from one individual to another [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020]. Building on this and the discussion in @sec-discussion_RA1, researchers could investigate how the selection of judges influences the formation of a "shared consensus" and whether these individual characteristics introduce systematic biases or distortions in the observed trait distribution [@Deffner_et_al_2022]. Furthermore, if such characteristics indeed undermine the "sample-freeness" assumption, then it becomes essential to explore strategies for controlling their influence. This includes investigating how many judges--and how many judgments per judge--are required to produce reliable trait estimates, as well as whether *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times [@Lawson_2015], can improve judgment consistency and accuracy.

@@

<!-- 1. Challenges for the framework: background knowledge and research context -->
Despite its promise, this study presents several challenges that researchers must navigate --challenges common to many quantitative studies. These challenges fall into two main categories. The first concerns the background knowledge researchers need to apply the proposed framework effectively. The second involves the increased attention researchers must now devote to their specific CJ research context.

<!-- 2. Challenge 1: background knowledge -->
In terms of background knowledge, researchers face two challenges: mastering Causal inference and developing proficiency in Bayesian inference methods. One example illustrating the need to master Causal inference is our assumption that predictor variables are "relevant" to the research context--interpreting relevance as the inclusion of these variables in a "sufficient adjustment set" [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. However, we do not fully explore what this entails or its implications for model specification, estimation, and inference. As a result, researchers must engage deeply with these complex ideas--and with others, such as how SCMs encode functional and probabilistic information--to apply the framework rigorously. To assist this effort, we provide key references throughout the study to guide motivated readers toward a deeper understanding of these foundational concepts[^9].

[^9]: refer to footnote 1, 2, and the detailed online document referenced in the Declarations section

Bayesian inference presents a second challenge. Researchers must acquire sufficient expertise to translate context-specific conceptual models into bespoke statistical models that meaningfully address their research questions. Although such translations are not unique to Bayesian approaches, we argue that Bayesian methods offer a more accessible route, as they reduce the need for specialized knowledge in optimization theory and techniques--often required by frequentist methods--thereby lowering the technical barriers for many applied researchers.

<!-- 3. Challenge 2: research context -->
In terms of the attention researchers must now devote to their specific CJ research context, researchers face two challenges: ensuring the conceptual model is a *faithful* representation for the CJ system, and verifying whether this model can accurately estimate the intended estimand from real data.

<!-- See fine point 6.1 and 6.2 [@Hernan_et_al_2025, pp. 79-81] -->

<!-- we are confident that these causal structure is faithful to a design -->

<!-- Ideally, we would like to assess this property through an *identification analysis* [@Schuessler_et_al_2023], where researchers assess how accurate a model computes an estimand based solely on its (causal) assumptions, regardless of random variability. As one can imagine these types of analysis requires formal proofs, nevertheless, we argue that simulation-based methods can provide preliminary assessments akin to an identification analysis. -->
